\documentclass[a4paper]{article}

\usepackage{newvbtm}
\usepackage{times}
\usepackage{fancyhdr}

\input{Defs.tex}
\title{XMF Reference Manual}
\author{Ceteva Ltd}

\pagestyle{fancy}
\lhead{}
\chead{The XMF Reference Manual}
\rhead{\thepage}
\cfoot{(c) 2003 Ceteva Ltd.}

\begin{document}

\maketitle
\tableofcontents

\section{Introduction}

XMF is a general engine that provides technolgies underlying a range of modelling
methods and techniques. The basic XMF engine is a virtual machine that runs
XMF VM code. The basic VM code runs in terms of a minimal language which is
a subset of languages such as MOF. This basic language is self aware and is used
as a basis for language definition. XMF is provided with a fairly extensive language, XMOF,
which is both an instance and an extension of the basic machine language. XMF is
not limited to this language, however, other languages can be defined in terms 
of the basic language and then loaded, making XMF very flexible. XMOF is also self aware
and can be used as a basis for language engineering since it provides a parser,
compiler and an interface to the XMF VM through various packages.

This document is the reference manual for the Ceteva Modelling Framework XMF. It
defines XMOF in terms of a collection of packages. This manual is not a tutorial 
for XMF although many of the packages and classes are documented in terms of examples.
The manual has been mechanically generated from the XMF source code through the Doc 
package.

\section{Installing and Running XMF}

The XMF system is a collection of directories under a root directory XMFHOME. You
can install XMF anywhere. The directory XMFHOME/bin contains scripts that control
the various ways in which XMF can be executed. Each script takes a number of command
line arguments, the first of which is always the path XMFHOME. Typically a script 
will start one of the saved images in XMFHOME/Images. A saved image is a resumable
snapshot of an XMF session. You can create your own images at any stage during a
session. The scripts are:
\begin{description}
\item[xmf] This is used to run a very minimal XMF image. You must
supply this script with XMFHOME and a binary file containing XMF machine instructions.
The image is started, the machine instructions are loaded and executed.
\item[xmfe] This is used to run an XMF command interpreter. The evaluator image
contains packages to parse and interpret commands. You must supply this
script with XMFHOME. Optionally, you can supply this script with a binary file
which is loaded and run.
\item[xmfc] This is used to run the XMF compiler from the command line or to
start an interactive session with the XMF compiler. You must supply this
script with XMFHOME. Optionally you may supply this script with an XMF
source file. The file should have an '.xmf' extension, although you need
not supply the extension as part of the a command line argument. If you supply
a file then it is compiled and the compiler terminates. If you do not supply a file
then an XMF command interpreter is started and you can interactively compile files
using the Compiler package.
\item[xmft] This script starts an XMF image and runs all the tests in
XMFHOME/Test. The tests should be fairly self explanatory. You must supply this
script with XMFHOME.
\item[make] This is used to reconstruct the XMF images in XMFHOME/Images.
\end{description}

\input{body.tex}

\end{document}